A Sound Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2022-09-01 paper/paper.tex
Minor fixes

\paragraph{Syntax and semantics of modal logics}
To implement our library, we use a \emph{deep embedding} of
formulas, i.e. we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive form: Set :=
| Lit    : nat          -> form
| Neg    : form -> form
| Box    : form -> form
| Dia    : form -> form
| And    : form -> form -> form
| Or     : form -> form -> form
| Implies: form -> form -> form.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logics. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$.

@@

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
axioms that form a system for modal logics. In this way, our definition can be
used by several systems of modal logic.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop)
                           : theory -> form -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: form)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: form),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: form)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: form)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax}, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : form -> form -> axiom
  | ax2  : form -> form -> form  -> axiom
  (* some code omitted *)
  | axK  : form -> form -> axiom
  | axPos: form -> form -> axiom
  | axT  : form -> axiom
  | axB  : form -> axiom
  | axK4 : form -> axiom
  | axD  : form -> axiom
  | axK5 : form -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 consists encode axioms for propositional logics
and the last 7 are for different formulas taken as axioms by systems of modal logics.
Axioms are built using the function \mintinline{coq}{instantiate}, which
receives an axiom and builds its corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): modalFormula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

>>>>

\paragraph{Syntax and semantics of modal logics}
To implement our library, we use a \emph{deep embedding} of
formulas, i.e. we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive form: Set :=
| Lit    : nat          -> form
| Neg    : form -> form
| Box    : form -> form
| Dia    : form -> form
| And    : form -> form -> form
| Or     : form -> form -> form
| Implies: form -> form -> form.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logics. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$. Furthermore, the type \mintinline{coq}{theory} is defined as a list
of formulas.

\begin{minted}{coq}
Definition theory := list form.
\end{minted}

@@

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
the set of axioms that form a system for some modal logic. In this way, our definition can be
used by several systems alike.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop)
                           : theory -> form -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: form)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: form),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: form)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: form)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax}, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : form -> form -> axiom
  | ax2  : form -> form -> form  -> axiom
  (* some code omitted *)
  | axK  : form -> form -> axiom
  | axPos: form -> form -> axiom
  | axT  : form -> axiom
  | axB  : form -> axiom
  | axK4 : form -> axiom
  | axD  : form -> axiom
  | axK5 : form -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 consists encode axioms for propositional logics
and the last 7 are for different formulas taken as axioms by systems of modal logics.
Axioms are built using the function \mintinline{coq}{instantiate}, which
receives an axiom and builds its corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): form :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.
